---
title: "COVID-19 Analytics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 8 
    fig_height: 6 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r import_dependecies, include=FALSE}
options(max.print = 1e3, scipen = 999, width = 1e2)
options(stringsAsFactors = F)

suppressPackageStartupMessages({
  # data manipulations
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(magrittr)
  
  # convert and formatting
  library(stringr)
  library(lubridate)
  
  # tools
  library(skimr)
  
  # graphics
  library(ggplot2)
  library(ggthemes)
})
```


```{r set_params, include=FALSE}
theme_set(theme_minimal())

lab_caption <- paste0(
    "Data source: Novel Coronavirus (COVID-19) Cases provided by Johns Hopkins University Center for Systems Science. \n",
    sprintf("Last updated: %s. ", format(Sys.time(), '%d %B, %Y')),
    "Source code: github.com/codez0mb1e/covid-2019"
  )            
```


## Load dataset

Get list of files in datasets container:

```{r}
input_data_container <- "../input/novel-corona-virus-2019-dataset.zip"

print(
  as.character(unzip(input_data_container, list = T)$Name)
)
```

Load `covid_19_data.csv` dataset:
```{r}
covid_data <- read.csv(unz(input_data_container, "covid_19_data.csv"), 
                       na.strings = c("NA", "None", ""),
                       header = T, sep = ",")

covid_data %>% sample_n(100) %>% as_tibble
```

Get dataset structure:
```{r}
covid_data %>% skim
```



## Preprocessing data

Set `area` column, processing `province_state` columns, and format dates columns:
```{r}
# repair names
names(covid_data) <- names(covid_data) %>% str_replace_all(fixed("."), "_") %>% str_to_lower

covid_data %<>% 
  rename(observation_date = observationdate) %>% 
  mutate(
    ## location processing
    province_state = str_trim(province_state),
    area = as.factor(
      case_when(
        province_state == "Hubei" ~ "Hubei",
        country_region == "US" ~ "US",
        str_detect(country_region, "China") ~ "China (exclude Hubei)",
        TRUE ~ "Rest of World")),
    
    ## dates processing
    observation_date = mdy(observation_date),
    last_update = parse_date_time(str_replace_all(last_update, "T", " "), 
                                  orders = c("%Y-%m-%d %H:%M:%S", "m/d/y %H:%M"))
  )
  

covid_data %>% 
  select(area, country_region, province_state, observation_date) %>% 
  arrange(desc(observation_date)) %>% 
  as_tibble
```

Get dataset structure after preprocessing:
```{r}
covid_data %>% skim
```



## COVID-19 spread

Get virus spread statistics grouped by `area`:

### Prepare data

Calculate total infected, recovered, and fatal cases:

```{r}
spread_df <- covid_data %>% 
  group_by(
    area, observation_date
  ) %>% 
  summarise(
    confirmed_total = sum(confirmed),
    deaths_total = sum(deaths),
    recovered_total = sum(recovered)
  )

spread_df %>% arrange(desc(observation_date))
```


### Visualize

Wordwide virus spread statistics:
```{r}
spread_df %>% 
  group_by(observation_date) %>% 
  summarise_at(vars(ends_with("_total")), sum) %>% 
  arrange(desc(observation_date))
```

```{r}

ggplot(
  spread_df %>% 
    ungroup %>% 
    mutate(confirmed_total = confirmed_total - recovered_total, deaths_total = -deaths_total) %>% 
    gather(key = "case_state", value = "cases_n", -c(area, observation_date))
  ) +
  
  geom_col(aes(x = observation_date, y = cases_n, fill = case_state), alpha = .75) +
  
  scale_x_date(date_labels = "%d %b", date_breaks = "7 days") +
  scale_fill_manual(element_blank(), 
                    labels = c("confirmed_total" = "Infected cases", "recovered_total" = "Recovered cases", "deaths_total" = "Fatal cases"),
                    values = c("confirmed_total" = "grey", "recovered_total" = "gold", "deaths_total" = "black")) +
  
  labs(x = "", y = "Number of cases", 
       title = "COVID-19 Spread", 
       subtitle = "Worldwide spread of the virus over time", 
       caption = lab_caption) +

  theme(
    legend.position = "top",
    plot.caption = element_text(size = 8)
  )

```


```{r}
ggplot(
    spread_df %>% 
      ungroup %>% 
      gather(key = "case_state", value = "cases_n", -c(area, observation_date))
  ) +
  
  geom_col(aes(x = observation_date, y = cases_n, fill = case_state), position = "fill", alpha = .75) +
  
  scale_x_date(date_labels = "%d %b", date_breaks = "7 days") +
  scale_fill_manual(element_blank(), 
                    labels = c("confirmed_total" = "Infected cases", "deaths_total" = "Fatal cases", "recovered_total" = "Recovered cases"),
                    values = c("confirmed_total" = "grey", "deaths_total" = "black", "recovered_total" = "gold")) +
  
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +

  labs(x = "", y = "% from the overall number of cases", 
       title = "COVID-19 Cases Structure", 
       subtitle = "Ratio of different types of cases", 
       caption = lab_caption) +
  
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 8)
  )

```



## COVID-19 daily spread

Get daily dynamics of new infected and recovered cases.

### Prepare data

```{r}
covid_daily <- spread_df %>% 
  ## calc cases per day stats
  mutate_at(
    vars(ends_with("_total")),
    list("per_day" = ~ (. - lag(.)))
  ) %>% 
  ungroup() %>% 
  ## calс cases per day 
  mutate_at(
    vars(ends_with("_per_day")), 
    list(~ replace_na(., 0))
  )
  
covid_daily %>% 
  filter(area == "Hubei") %>% 
  arrange(desc(observation_date))
```


### Visualize

```{r}
ggplot(
    covid_daily %>% 
      transmute(
        area, observation_date,
        confirmed_total_per_day,
        increase_per_day = confirmed_total_per_day - (recovered_total_per_day + deaths_total_per_day),
        deaths_total_per_day = -deaths_total_per_day,
        recovered_total_per_day = -recovered_total_per_day
      ) %>% 
      gather(
        key = "case_state", value = "cases_n", -c(area, observation_date, increase_per_day)
      ), 
    aes(x = observation_date)
  ) +
  
  geom_col(aes(x = observation_date, y = cases_n, fill = case_state), alpha = .75) +
  geom_smooth(aes(y = increase_per_day), method = "loess", color = "grey", alpha = .25) +
  
  scale_x_date(date_labels = "%d %b", date_breaks = "7 days") +
  scale_fill_manual(element_blank(), 
                    labels = c("confirmed_total_per_day" = "Infected cases", "recovered_total_per_day" = "Recovered cases", "deaths_total_per_day" = "Fatal cases"),
                    values = c("confirmed_total_per_day" = "grey", "recovered_total_per_day" = "gold", "deaths_total_per_day" = "black")) +
  
  facet_grid(area ~ ., scales = "free") +
  
  labs(title = "COVID-19 Daily Spread", 
       subtitle = "Daily dynamics by Hubei, China (exclude Hubei), US, and Rest of World. \nLines - infected cases minus recovered and fatal. Negative values indicate recovered and fatal cases", 
       x = "", y = "Number of cases per day",
       caption = lab_caption) +
  
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 8)
  )
```


## COVID-19 mortality rate

### Prepare data

```{r}
mortality_df <- covid_data %>% 
  group_by(area, observation_date) %>% 
  summarise(
    confirmed_total = sum(confirmed),
    deaths_total = sum(deaths),
    recovered_total = sum(recovered)
  ) %>% 
  ungroup() %>% 
  inner_join(
    covid_data %>% 
      filter(deaths > 10) %>% 
      group_by(area) %>% 
      summarise(reference_date = min(observation_date)),
    by = "area"
  ) %>% 
  mutate(
    confirmed_deaths_rate = deaths_total/confirmed_total,
    recovered_deaths_rate = deaths_total/(recovered_total + deaths_total),
    n_days = observation_date %>% difftime(reference_date, units = "days") %>% as.numeric
  ) %>% 
  filter(n_days >= 0)


mortality_df %>% 
  filter(area == "US") %>% 
  arrange(desc(observation_date)) %>% 
  select(area, ends_with("_date"), recovered_total, contains("deaths_"))
```


### Visualize

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  
  geom_area(aes(y = recovered_deaths_rate), alpha = .75, fill = "gold") +
  geom_area(aes(y = confirmed_deaths_rate), alpha = .75, fill = "black") +
  
  scale_y_continuous(labels = scales::percent_format(accuracy = .1)) +
  
  facet_grid(area ~ ., scales = "free") +
  
  labs(x = "Number of days since 10th fatal case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Rate by Hubei, China (exclude Hubei), US, and Rest of World. \nGold area - fatal to recovered cases ratio, grey area - fatal to infected cases ratio.", 
       caption = lab_caption) +
  
  theme(
    legend.position = "top",
    plot.caption = element_text(size = 8)
  )
```

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  
  geom_hline(aes(yintercept = mean(mortality_df$confirmed_deaths_rate)), linetype = "dashed", color = "black", alpha = .33) +
  annotate(geom = "text", label = "Mean mortality rate (over time)", x = 8, y = mean(mortality_df$confirmed_deaths_rate), vjust = -1) +
  geom_line(aes(y = confirmed_deaths_rate, color = area), size = 1, alpha = .75) +

  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(x = "Number of days since 10th fatal case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Fatal to infected cases ratio by Hubei, China (without Hubei), US, and Rest of World", 
       caption = lab_caption) +
  
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 8)
  )

```

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  
  geom_hline(aes(yintercept = mean(mortality_df$recovered_deaths_rate)), linetype = "dashed", color = "black", alpha = .33) +
  annotate(geom = "text", label = "Mean mortality rate (over time)", x = 8, y = mean(mortality_df$recovered_deaths_rate), vjust = -1) +
  geom_line(aes(y = recovered_deaths_rate, color = area), size = 1, alpha = .75) +

  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(x = "Number of days since 10th fatal case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Fatal to recovered cases ratio by Hubei, China (exclude Hubei), US, and Rest of World", 
       caption = lab_caption) +
  
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.caption = element_text(size = 8)
  )
```


## Enrich COVID dataset with world population

### Load world population

Get datasets list:

```{r}
world_population_data_container <- "../input/world-population-19602018.zip"
print(
  as.character(unzip(world_population_data_container, list = T)$Name)
)
```

Load world population dataset:

```{r}
world_population <- read.csv(unz(world_population_data_container, "population_clean.csv"),
                             header = T, sep = ",")

world_population %>% skim_to_wide
```

Select relevant columns:

```{r}
world_population_2018 <- world_population %>%
  transmute(
    country = Country.Name,
    n = X2018
  )

world_population_2018 %>% arrange(-n) %>% as_tibble
```


### Preprocessing

Get unmatched countries:

```{r}
not_matched_countries <- setdiff(
  unique(covid_data$country_region),
  unique(world_population_2018$country)
)

covid_data %>% 
  filter(country_region %in% not_matched_countries) %>% 
  group_by(country_region) %>% 
  summarise(n = sum(confirmed)) %>% 
  filter(n > 1000) %>% 
  arrange(-n)
```

Correct top of unmached countries.

```{r}
world_population_2018[world_population_2018$country == "China", ]$country <- "Mainland China"
world_population_2018[world_population_2018$country == "Iran, Islamic Rep.", ]$country <- "Iran"
world_population_2018[world_population_2018$country == "Korea, Rep.", ]$country <- "South Korea"
world_population_2018[world_population_2018$country == "United States", ]$country <- "US"
world_population_2018[world_population_2018$country == "United Kingdom", ]$country <- "UK"
world_population_2018[world_population_2018$country == "Hong Kong SAR, China", ]$country <- "Hong Kong"
world_population_2018[world_population_2018$country == "Egypt, Arab Rep.", ]$country <- "Egypt"
world_population_2018[world_population_2018$country == "Russian Federation", ]$country <- "Russia"
```


And updated matching:

```{r}
not_matched_countries <- setdiff(
  unique(covid_data$country_region),
  unique(world_population_2018$country)
)

covid_data %>% 
  filter(country_region %in% not_matched_countries) %>% 
  group_by(country_region) %>% 
  summarise(n = sum(confirmed)) %>% 
  filter(n > 1000) %>% 
  arrange(-n)
```

Much better :)



### Infected, recovered, fatal, and active cases

Calculate number of infected, recovered, fatal, and active cases grouped by country:

```{r}
countries_cases <- covid_data %>% 
  ## calc number of infected, recovered, and fatal cases by country
  group_by(country_region, observation_date) %>% 
  summarise(
    confirmed_total = sum(confirmed),
    recovered_total = sum(recovered),
    deaths_total = sum(deaths)
  ) %>% 
  mutate(
    active_total = confirmed_total - (recovered_total + deaths_total)
  ) %>% 
  ungroup() %>% 
  
  ## calc number of days since first infected case
  inner_join(
    covid_data %>% 
      group_by(country_region) %>% 
      filter(confirmed > 0) %>% 
      summarise(first_confirmed_date = min(observation_date)),
    by = "country_region"
  ) %>%
  mutate(
    n_days_since_1st_confirmed = observation_date %>% difftime(first_confirmed_date, units = "days") %>% as.numeric
  ) %>% 
  filter(n_days_since_1st_confirmed >= 0) %>% 
  
  ## calc number of days since 10th fatal case
  left_join(
    covid_data %>% 
      group_by(country_region, observation_date) %>% 
      summarise(deaths_total = sum(deaths)) %>% 
      filter(deaths_total > 0) %>% 
      summarise(first_deaths_case_date = min(observation_date)),
    by = "country_region"
  ) %>% 
  mutate(
    n_days_since_1st_deaths = observation_date %>% difftime(first_deaths_case_date, units = "days") %>% as.numeric,
    n_days_since_1st_deaths = if_else(n_days_since_1st_deaths >= 0, n_days_since_1st_deaths, NA_real_)
  )
```

View statistics in US:

```{r}
countries_cases %>% 
  filter(country_region == "US") %>% 
  arrange(-n_days_since_1st_confirmed)
```

View statistics in Russia:

```{r}
countries_cases %>% 
  filter(country_region == "Russia") %>% 
  arrange(-n_days_since_1st_confirmed)

```



### Join COVID-19 dataset with world population

```{r}
countries_cases %<>% 
  ## join w/ population dataset
  inner_join(
    world_population_2018 %>% rename(population_n = n), 
    by = c("country_region" = "country") 
  ) %>% 
  
  ## filter only countries with population greater than 1M
  filter(population_n > 1e6) %>% 
  
  ## calculate count of cases per 1M population
  mutate_at(
    vars(ends_with("_total")),
    list("per_1M" = ~ ./population_n*1e6)
  )
  
countries_cases %>% 
  filter(country_region == "Russia") %>% 
  arrange(-n_days_since_1st_confirmed) %>% 
  select(country_region, n_days_since_1st_confirmed, population_n, starts_with("confirmed_"))
```


### TOPs countries by infected, active, and fatal cases

Calculate countries stats whose populations were most affected by the virus:

```{r}
countries_cases_last <- countries_cases %>% 
  ## get last observation for country
  group_by(country_region) %>% 
  arrange(desc(observation_date)) %>% 
  filter(row_number() == 1) %>% 
  ungroup() %>% 
  ## filter countries with infected cases greatet than 1K
  filter(confirmed_total > 1000)
```



#### Top countries by infected cases 

```{r}
countries_cases_last %>% 
  arrange(-confirmed_total_per_1M) %>% 
  select(
    country_region, population_n,
    confirmed_total, confirmed_total_per_1M,
    n_days_since_1st_confirmed
  )
```


#### Top countries by active cases 

```{r}
countries_cases_last %>% 
  arrange(-active_total_per_1M) %>% 
  select(
    country_region, population_n,
    active_total, active_total_per_1M,
    n_days_since_1st_confirmed
  )
```


#### Top countries by fatal cases 

```{r}
countries_cases_last %>% 
  arrange(-deaths_total_per_1M) %>% 
  select(
    country_region, population_n,
    deaths_total, deaths_total_per_1M,
    n_days_since_1st_confirmed
  )
```



#### Select countries to monitoring

Get top N countries:

```{r}
top_n <- 10

observed_countries <- countries_cases_last %>% 
  arrange(desc(active_total_per_1M)) %>% 
  top_n(top_n, active_total_per_1M) %>% 
  select(country_region) %>% 
  as_vector %>% sort


# Add countries that we are interested in
observed_countries <- c(observed_countries, "US", "Mainland China", "South Korea","Iran") %>% unique
observed_countries 
```


### Active cases per 1M population vs number of days since 1st infected case

```{r warning=FALSE}
confirmed_cases_last <- countries_cases %>% 
  group_by(country_region) %>% 
  arrange(desc(observation_date)) %>% 
  filter(row_number() == 1) %>% 
  ungroup() %>% 
  transmute(
    country_region, 
    active_total_per_1M_last = active_total_per_1M,
    n_days_since_1st_confirmed
  )

active_case_since_1st_infected <- countries_cases %>% 
  filter(country_region %in% observed_countries) %>% 
  left_join(
    confirmed_cases_last,
    by = c("country_region", "n_days_since_1st_confirmed")
  ) %>% 
  mutate(
    double_every_7d = (1 + 1/7)^n_days_since_1st_confirmed, # double every week
    double_every_3d = (1 + 1/3)^n_days_since_1st_confirmed, # double every 3 days
    double_every_2d = (1 + 1/2)^n_days_since_1st_confirmed, # double every 2 days
    double_every_1d = (1 + 1)^n_days_since_1st_confirmed  # double every day
  )


ggplot(active_case_since_1st_infected, aes(x = n_days_since_1st_confirmed)) +
  
  geom_line(aes(y = double_every_7d), linetype = "dashed", color = "red", alpha = .45) +
  geom_line(aes(y = double_every_3d), linetype = "dashed", color = "red", alpha = .6) + 
  geom_line(aes(y = double_every_2d), linetype = "dashed", color = "red", alpha = .85) + 
  
  geom_line(aes(y = active_total_per_1M, color = country_region), show.legend = F) +
  geom_text(aes(y = active_total_per_1M_last + 20, label = country_region, color = factor(country_region)),
            hjust = 0.5, vjust = 0, check_overlap = T, show.legend = F, fontface = "bold", size = 3.6) +
  
  annotate(geom = "text", label = "Active cases double \n every 2 days", x = 17, y = 1550, vjust = 0, size = 3.2) +
  annotate(geom = "text", label = "...every 3 days", x = 25, y = 1400, vjust = 0, size = 3.2) +
  annotate(geom = "text", label = "...every week", x = 55, y = 1600, vjust = 0, size = 3.2) +
  
  ylim(c(0, max(countries_cases$active_total_per_1M) + 100)) +
  
  labs(x = "Number of days since 1st infected case", y = "Active cases per 1M population", 
       title = "COVID-19 Аctive Сases", 
       subtitle = "Number of active cases per 1 million population since 1st infected case", 
       caption = lab_caption) +

  theme(
    plot.caption = element_text(size = 8)
  )
```


### Active cases per 1 million population vs number of days since 1st fatal case

```{r warning=FALSE}
deaths_cases_last <- countries_cases %>% 
  group_by(country_region) %>% 
  arrange(desc(observation_date)) %>% 
  filter(row_number() == 1) %>% 
  ungroup() %>% 
  transmute(
    country_region, 
    active_total_per_1M_last = active_total_per_1M,
    n_days_since_1st_deaths
  )


ggplot(
  countries_cases %>% 
    filter(country_region %in% observed_countries) %>% 
    left_join(
      deaths_cases_last,
      by = c("country_region", "n_days_since_1st_deaths")
    ) %>% 
    mutate(
      double_every_7d = (1 + 1/7)^n_days_since_1st_deaths, # double every week
      double_every_3d = (1 + 1/3)^n_days_since_1st_deaths, # double every 3 days
      double_every_2d = (1 + 1/2)^n_days_since_1st_deaths, # double every 2 days
      double_every_1d = (1 + 1)^n_days_since_1st_deaths  # double every day
    ),
    aes(x = n_days_since_1st_deaths)
  ) +
  
  geom_line(aes(y = double_every_7d), linetype = "dashed", color = "red", alpha = .45) +
  geom_line(aes(y = double_every_3d), linetype = "dashed", color = "red", alpha = .6) + 
  geom_line(aes(y = double_every_2d), linetype = "dashed", color = "red", alpha = .85) + 
  geom_line(aes(y = double_every_1d), linetype = "dashed", color = "red", alpha = 1) +
  
  geom_line(aes(y = active_total_per_1M, color = country_region)) +
  geom_text(aes(y = active_total_per_1M_last + 20, label = country_region, color = factor(country_region)), 
            hjust = 0.5, vjust = 0, check_overlap = T, show.legend = F, fontface = "bold", size = 3.6) +
  
  annotate(geom = "text", label = "Active cases double \n every day", x = 10, y = 1050, vjust = 0, size = 3.2) +
  annotate(geom = "text", label = "...every 2 days", x = 18, y = 1500, vjust = 0, size = 3.2) +
  annotate(geom = "text", label = "...every 3 days", x = 22, y = 1250, vjust = 0, size = 3.2) +
  annotate(geom = "text", label = "...every week", x = 200, y = 35, vjust = 0, size = 3.2) +
  
  xlim(c(0, 42)) +
  ylim(c(0, max(countries_cases$active_total_per_1M) + 100)) +
  
  labs(x = "Number of days since 1st fatal case", y = "Active cases per 1M population", 
       title = "COVID-19 Аctive Сases", 
       subtitle = "Number of active cases per 1 million population since 1st fatal case", 
       caption = lab_caption) +
  
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.caption = element_text(size = 8)
  )
```
 


