---
title: "COVID-19 Analytics"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 12 
    fig_height: 8 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r import_dependecies, include=FALSE}
options(max.print = 1e3, scipen = 999, width = 1e2)
options(stringsAsFactors = F)

suppressPackageStartupMessages({
  # data manipulations
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(magrittr)
  
  # convert and formatting
  library(stringr)
  library(lubridate)
  
  # tools
  library(skimr)
  
  # graphics
  library(ggplot2)
  library(ggthemes)
})
```


```{r set_params, include=FALSE}
caption <- "Data source: Novel Corona Virus 2019 Dataset. \n© 2020, Dmitry Petukhov. CC BY-SA 4.0 license."
```


## Load dataset

Get list of files in datasets container:

```{r}
input_data_container <- "../input/novel-corona-virus-2019-dataset.zip"

print(
  as.character(unzip(input_data_container, list = T)$Name)
)
```

Load `covid_19_data.csv` dataset:
```{r}
covid_data <- read.csv(unz(input_data_container, "covid_19_data.csv"), 
                       na.strings = c("NA", "None", ""),
                       header = T, sep = ",")

covid_data %>% sample_n(100) %>% as_tibble
```

Get dataset structure:
```{r}
str(covid_data)
```



## Preprocessing data

Set `area` column, processing `province_state` columns, and format dates columns:
```{r}
# repair names
names(covid_data) <- names(covid_data) %>% str_replace_all(fixed("."), "_") %>% str_to_lower

covid_data %<>% 
  rename(observation_date = observationdate) %>% 
  mutate(
    ## location processing
    province_state = str_trim(province_state),
    area = as.factor(
      case_when(
        province_state == "Hubei" ~ "Hubei",
        country_region == "US" ~ "US",
        str_detect(country_region, "China") ~ "China without Hubei",
        TRUE ~ "Rest of World")),
    
    ## dates processing
    observation_date = mdy(observation_date),
    last_update = parse_date_time(str_replace_all(last_update, "T", " "), 
                                  orders = c("%Y-%m-%d %H:%M:%S", "m/d/y %H:%M"))
  )
  

covid_data %>% as_tibble
```

Get dataset structure after preprocessing:
```{r}
covid_data %>% skim_to_wide
```



## COVID-19 spread

Get virus spread statistics grouped by `area`:

### Prepare data

Calculate total confirmed, deaths, and recovered cases:

```{r}
spread_df <- covid_data %>% 
  group_by(
    area, observation_date
  ) %>% 
  summarise(
    confirmed_total = sum(confirmed),
    deaths_total = sum(deaths),
    recovered_total = sum(recovered)
  ) %>% 
  mutate(
    active_total = confirmed_total - (recovered_total + deaths_total)
  )

spread_df %>% arrange(desc(observation_date))
```


### Visualize

```{r}
# TODO: add color legend

ggplot(spread_df, aes(observation_date)) +
  
  geom_col(aes(y = confirmed_total), alpha = .75, fill = "grey") +
  geom_col(aes(y = recovered_total), alpha = .75, fill = "gold") +
  geom_col(aes(y = -deaths_total), alpha = .75, fill = "black") +

  scale_x_date(date_labels = "%d %b", date_breaks = "7 days") +
  scale_color_discrete(name = "Infected cases") +

  labs(x = "", y = "Number of cases", 
       title = "COVID-19 Spread", 
       subtitle = "Virus spread by Hubei, China (without Hubei), US, and Rest of World. \n* Grey bar - all infected cases; gold bar - all recovered cases; black bar - all deaths cases.", 
       caption = caption) +
  
  theme_minimal() +
  theme(legend.position = "top")
```



## Dynamics of Infection

Get daily dynamics of new infected and recovered cases.

### Prepare data

```{r}
covid_daily <- spread_df %>% 
  ## calc cases per day stats
  mutate_at(
    vars(ends_with("_total")),
    list("per_day" = ~ (. - lag(.)))
  ) %>% 
  ungroup() %>% 
  ## calс infected and retired cases
  transmute(
    area, observation_date,
    retired_per_day = recovered_total_per_day + deaths_total_per_day,
    infected_per_day = confirmed_total_per_day
  ) %>% 
  mutate_at(
    vars(ends_with("_per_day")), 
    list(~ replace_na(., 0))
  )
  
covid_daily %>% 
  filter(area == "Hubei") %>% 
  arrange(desc(observation_date))
```


### Visualize

```{r}
ggplot(covid_daily, aes(x = observation_date)) +
  
  geom_col(aes(y = -retired_per_day), fill = "gold", alpha = .7) +
  geom_col(aes(y = infected_per_day), fill = "black", alpha = .7) +
  geom_smooth(aes(y = infected_per_day - retired_per_day), method = "loess", color = "grey", alpha = .25) +
  
  facet_grid(area ~ ., scale = "free") +
  
  labs(title = "COVID-19 New Cases", 
       subtitle = "Daily dynamics by Hubei, China (without Hubei), US, and Rest of World. \n* Lines - active (infected minus healed and deaths) cases; gold bar - healed plus deaths cases; black bar - infected cases.", 
       x = "", y = "Number of cases per day",
       caption = caption) +
  
  theme_minimal()
```


## Mortality rate

### Prepare data

```{r}
mortality_df <- covid_data %>% 
  group_by(area, observation_date) %>% 
  summarise(
    confirmed_total = sum(confirmed),
    deaths_total = sum(deaths),
    recovered_total = sum(recovered)
  ) %>% 
  ungroup() %>% 
  inner_join(
    covid_data %>% 
      filter(deaths > 10) %>% 
      group_by(area) %>% 
      summarise(reference_date = min(observation_date)),
    by = "area"
  ) %>% 
  mutate(
    confirmed_deaths_rate = deaths_total/confirmed_total,
    recovered_deaths_rate = deaths_total/(recovered_total + deaths_total),
    n_days = observation_date %>% difftime(reference_date, units = "days") %>% as.numeric
  ) %>% 
  filter(n_days >= 0)


mortality_df %>% 
  filter(area == "US") %>% 
  arrange(desc(observation_date)) %>% 
  select(area, ends_with("_date"), recovered_total, contains("deaths_"))
```


### Visualize

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  geom_area(aes(y = recovered_deaths_rate), alpha = .75, fill = "gold") +
  geom_area(aes(y = confirmed_deaths_rate), alpha = .75, fill = "black") +
  
  scale_y_continuous(labels = scales::percent_format(accuracy = .1)) +
  
  facet_grid(area ~ ., scales = "free") +
  
  labs(x = "Number of days since 10-th death case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Rate by Hubei, China (without Hubei), US, and Rest of World. \n* Gold area - deaths to recovered cases ratio, grey area - deaths to confirmed cases ratio.", 
       caption = caption) +
  
  theme_minimal()
```

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  
  geom_hline(aes(yintercept = mean(mortality_df$confirmed_deaths_rate)), linetype = "dashed", color = "black", alpha = .33) +
  annotate(geom = "text", label = "Mean mortality rate (all time)", x = 4, y = mean(mortality_df$confirmed_deaths_rate), vjust = -1) +
  geom_line(aes(y = confirmed_deaths_rate, color = area), size = 1, alpha = .75) +

  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(x = "Number of days since 10-th death case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Deaths to confirmed cases ratio by Hubei, China (without Hubei), US, and Rest of World", 
       caption = caption) +
  
  theme_minimal() +
  theme(legend.position = "top", 
        legend.title = element_blank())

```

```{r}
ggplot(mortality_df, aes(x = n_days)) +
  
  geom_hline(aes(yintercept = mean(mortality_df$recovered_deaths_rate)), linetype = "dashed", color = "black", alpha = .33) +
  annotate(geom = "text", label = "Mean mortality rate (all time)", x = 4, y = mean(mortality_df$recovered_deaths_rate), vjust = -1) +
  geom_line(aes(y = recovered_deaths_rate, color = area), size = 1, alpha = .75) +

  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  
  labs(x = "Number of days since 10-th death case", y = "Mortality rate", 
       title = "COVID-19 Mortality Rate", 
       subtitle = "Deaths to recovered cases ratio by Hubei, China (without Hubei), US, and Rest of World", 
       caption = caption) +
  
  theme_minimal() +
  theme(legend.position = "top", 
        legend.title = element_blank())
```


